### 2016-10-29 latex/input/experiments.tex
incorporated most of reviewer suggestions in paper

\begin{table}[h]
\begin{center} 
%\hfill
\caption{Performance with even train/test split.} 
\label{tab:even-comp}
\begin{tabular}{|c|ccc|}\hline
Method & Accuracy & F1-Score & AUC-ROC\\\hline
MISVM & --- & --- & ---\\\hline
sbMIL & --- & --- & ---\\\hline
GICF & 63.04 & 68.33 & 66.19\\\hline %59.24,63.05,66.19
CAMIL D-BoW & 86.34 & 87.18 & 95.93\\\hline
CAMIL H-BoW & \bf{90.71} & \bf{89.70} & \bf{97.63}\\\hline
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\begin{center}
\caption{Classification time and memory usage with even train/test split.} 
\label{tab:time-comp}
\begin{tabular}{|c|cc|}\hline
Method & Classification Time & Memory Usage\\\hline
MISVM & --- & Memory Error\\\hline
sbMIL & --- & Memory Error\\\hline
GICF & 8 hours, 44 mins, 27 secs & 2.646 GB\\\hline
CAMIL D-BoW & \bf{6 minutes, 58 seconds} & \bf{545.293 MB}\\\hline
CAMIL H-BoW & 7 minutes, 56 seconds & 546.297 MB\\\hline
\end{tabular}
\end{center}
\end{table}

In the original MGWAS paper, the authors use 344 patients as a training set and 23 as a test set. However, when comparing to other MIL algorithms, we wished to have a more balanced training vs. test set split. We put 184 patients in the training set and 183 in the test set. Since this used a different training set than the one in the MGWAS paper, we do not compare our results to theirs here. Table \ref{tab:even-comp} shows the comparison of results between CAMIL, GICF, MISVM, and sbMIL, while Table \ref{tab:time-comp} compares the classification time and memory usage. CAMIL methods significantly outperform GICF, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant. CAMIL took significantly less time than GICF for two main reasons: (i) GICF requires representing each read as an array of length 64 (for k-mer length 3), while CAMIL reduces the data size with clustering and feature extraction; (ii) GICF requires expensive pairwise comparisons between each pair of instances in a mini-batch. MISVM and sbMIL require computing a kernel matrix of size N*N, where N = the number of instances. Since this dataset involved millions of instances, these methods crashed with memory errors. This is why they are shown as ``---" in the results tables.

@@

We wanted to compare MISVM and sbMIL to GICF and CAMIL, so we used a subset of the instances for each patient so that MISVM and sbMIL would not crash. They could only be run on 0.1\% of the reads for each patient, and even then took over 32.5 GB of memory. The results are shown in Table \ref{tab:subset-comp}. MISVM and sbMIL also only achieved 50.8\% accuracy, only slightly better than a random guess. The F1 score could not be computed because there were no True Positives or False Positives. GICF and CAMIL, while experiencing a performance drop due to the massive information loss, still performed much better, with the CAMIL methods outperforming GICF again. CAMIL D-BoW outperformed CAMIL H-BoW this time, because the distance calculations are less affected by having fewer reads than the histogram method.

MISVM and sbMIL performed the worst overall. This makes sense, as they make the standard MIL assumption, which doesn't make sense in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens. Additionally, they are instance space methods that do not leverage bag-level information. The performance of these two methods serves to illustrate why many of the classic MIL algorithms with standard assumptions will not be effective in this domain. GICF performs better than MISVM and sbMIL, which makes sense given the fact that it follows the collective assumption. It also has the benefit of calculating instance labels, which we explore further in the next section. However, GICF is still an instance space method, so it makes sense that CAMIL outperformed it.

@@

Table \ref{tab:test-comp} compares CAMIL to the method used by Qin et al., mRMR + SVM. We initially tested CAMIL on the same 23 patient test set that Qin et al. used, for which CAMIL H-BoW had 100\% accuracy and AUC, while mRMR + SVM had 0.81 AUC as reported by Qin et al. \cite{qin041012}. We know CAMIL is not 100\% accurate, so we validated it by averaging the results of 10 independent trials in which we selected 23 patients randomly out of the 367 to serve as the test set, with the other 344 of the training set. Table \ref{tab:test-comp} shows that CAMIL significantly outperformed mRMR + SVM on these trials, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant.

>>>>

\begin{table}[h]
\begin{center} 
%\hfill
\caption{Performance with even train/test split.} 
\label{tab:even-comp}
\begin{tabular}{|c|ccccc|}\hline
Method & Accuracy & F1-Score & AUC-ROC & CV Acc. & CV F1\\\hline
MISVM & --- & --- & --- & --- & ---\\\hline
sbMIL & --- & --- & --- & --- & ---\\\hline
GICF & 63.04 & 68.33 & 66.19 & --- & ---\\\hline %59.24,63.05,66.19
GICF-Cluster & 79.31 & 82.35 & 82.38 & --- & ---\\\hline
CAMIL D-BoW & 86.34 & 87.18 & 95.93 & 82.07 & 83.07\\\hline
CAMIL H-BoW & \bf{90.71} & \bf{89.70} & \bf{97.63} & \bf{89.13} & \bf{88.09}\\\hline
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\begin{center}
\caption{Classification time and memory usage with even train/test split.} 
\label{tab:time-comp}
\begin{tabular}{|c|cc|}\hline
Method & Classification Time & Memory Usage\\\hline
MISVM & --- & Memory Error\\\hline
sbMIL & --- & Memory Error\\\hline
GICF & 8 hours, 44 mins, 27 secs & 2.646 GB\\\hline
GICF-Cluster & 24 minutes, 51 secs & \bf{500.07 MB}\\\hline 
CAMIL D-BoW & 5 minutes, 33 seconds & 545.293 MB\\\hline
CAMIL H-BoW & \bf{5 minutes, 25 seconds} & 546.297 MB\\\hline
\end{tabular}
\end{center}
\end{table}

Qin et al. use 344 patients as a training set and 23 as a test set. However, when comparing to other MIL algorithms, we wished to have a more balanced training vs. test set split. We put 184 patients in the training set and 183 in the test set. Since this used a different training set than the one in the MGWAS paper, we do not compare our results to theirs here. Table \ref{tab:even-comp} shows the comparison of results between CAMIL, GICF, MISVM, and sbMIL, while Table \ref{tab:time-comp} compares the classification time and memory usage. CAMIL methods significantly outperform GICF, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant. We attempted to improve on GICF's results by selecting only one read from each cluster for each patient and discarding the other reads.  This significantly reduced the computation time and improved results, showing how important the clustering step is. These results are listed as ``GICF-Cluster" in the tables above. Even with this step, CAMIL was faster and more effective than GICF-Cluster, demonstrating the effectiveness of our feature extraction method. We performed Leave One Out Cross Validation for the Accuracy and F1-Score metrics (denoted in the table as CV Acc. and CV F1, respectively). CAMIL's cross validation results were slightly worse than its test set results but still significantly outperformed GICF. Cross validation results for GICF and GICF-Cluster were infeasible to calculate due to extremely long computaion time. CAMIL took much less time than GICF for two main reasons: (i) GICF requires representing each read as an array of length 64 (for k-mer length 3), while CAMIL reduces the data size with clustering and feature extraction; (ii) GICF requires expensive pairwise comparisons between each pair of instances in a mini-batch. MISVM and sbMIL require computing a kernel matrix of size N*N, where N = number of instances. Since this dataset had millions of instances, these methods crashed with memory errors, and are shown as ``---" in the tables. We attempted to resolve these issues by applying the same methods that we used for GICF-Cluster, but still received memory errors.

@@

We wanted to compare MISVM and sbMIL to GICF and CAMIL, so we used a subset of the instances for each patient so that MISVM and sbMIL would not crash. They could only be run on 0.1\% of the reads for each patient, and even then took over 32.5 GB of memory. The results are shown in Table \ref{tab:subset-comp}. MISVM and sbMIL only achieved 50.8\% accuracy. The F1 score could not be computed because there were no True or False Positives. GICF and CAMIL, while experiencing a performance drop due to the massive information loss, still performed much better, with the CAMIL methods outperforming GICF again. CAMIL D-BoW outperformed CAMIL H-BoW this time, because the distance calculations are less affected by having fewer reads than the histogram method.

MISVM and sbMIL performed the worst overall. This makes sense, as they make the standard MIL assumption, which is not helpful in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens. Additionally, they are instance space methods that do not leverage bag-level information. The performance of these two methods serves to illustrate why many of the classic MIL algorithms with standard assumptions will not be effective in this domain. GICF performs better than MISVM and sbMIL, which makes sense given the fact that it follows the collective assumption. It also has the benefit of calculating instance labels, which we explore further in the next section. However, GICF is still an instance space method, so it makes sense that CAMIL outperformed it.

@@

Table \ref{tab:test-comp} compares CAMIL to the method used by Qin et al., mRMR + SVM. We initially tested CAMIL on the same 23 patient test set that Qin et al. used, for which CAMIL H-BoW had 100\% accuracy and AUC, while mRMR + SVM had 0.81 AUC as reported by Qin et al. \cite{qin041012}. We know CAMIL is not 100\% accurate, so we validated it by averaging the results of 10 independent trials in which we selected 23 patients randomly out of the 367 to serve as the test set, with the other 344 of the training set. Table \ref{tab:test-comp} shows that CAMIL significantly outperformed mRMR + SVM on these trials, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant. Leave-one-out cross validation generally yielded similar results to the test set results, except precision was sometimes lower.

### 2016-10-29 latex/input/methods.tex
results section comparison to mRMR+SVM updated
incorporated most of reviewer suggestions in paper

We use UCLUST within our study, which is one of the most widely used and cited metagenome clustering methods and has been shown to be amongst the most effective in terms of speed and accuracy in benchmarking studies \cite{bonder090112, sun042711}. UCLUST seeks to ensure that, for some similarity T, the following conditions hold: (i) all cluster centroids have a similarity of less than T to each other; and (ii) all points in a cluster have a similarity of greater than T to the cluster centroid \cite{Edgar10}. Thus, each \emph{centroid} defines the center of a cluster, and the distance T defines the radius of the cluster, such that any point that has a similarity of greater than T to the centroid is within the radius and is thus part of the cluster. UCLUST is a heuristic algorithm that has several optimizations to improve speed, thus condition (i) above is not always guaranteed \cite{Edgar10}. UCLUST proceeds in a greedy, iterative manner. The first sequence in the input file becomes a new cluster centroid. For each new sequence in the file, it is compared with each of the existing cluster centroids in order. As soon as it is compared with a centroid that it has a similarity of greater than T with, it becomes part of that cluster. If the read is not similar enough with any of the existing cluster centroids, it becomes the centroid of a new cluster. The similarity measure T is defined as a string similarity between the two nucleotide sequences that counts the number of character placements that they have in common and then divides that number by the length of the reads, with terminal characters excluded \cite{Edgar10}.

@@

We used a "vocabulary-based" feature extraction method. An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}. Step (i) is covered by our assembly and clustering process, while step (iii) is covered by performing classification with a standard SVM classifier based on the extracted feature vectors. In this case, we used svm-light \cite{joachims08}. The choice of classifier is not very important for BoW methods \cite{amores13}. Below, we describe our feature selection methods for step (ii), which we implemented in Python, as well as the rationale for using these methods.

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{./instance-labels.png}
\caption{This diagram illustrates why static instance labels are not sufficient for phenotype prediction. A patient with 6 of the blue microbe or 6 of the green microbe may be healthy, while a patient with 3 of each is sick. Static instance labels cannot capture this relationship. This is also explained by Amores in his MIL taxonomy \cite{amores13}.} \label{instance-labels}
\end{figure}

>>>>

We use UCLUST within our study, which is one of the most widely used and cited metagenome clustering methods and has been shown to be amongst the most effective in terms of speed and accuracy in benchmarking studies \cite{bonder090112, sun042711}. UCLUST seeks to ensure that, for some similarity T, the following conditions hold: (i) all cluster centroids have a similarity of less than T to each other; and (ii) all points in a cluster have a similarity of greater than T to the cluster centroid \cite{Edgar10}. UCLUST proceeds in a greedy, iterative manner. The first sequence in the input file becomes a new cluster centroid. For each new sequence in the file, it is compared with each of the existing cluster centroids in order. As soon as it is compared with a centroid that it has a similarity of greater than T with, it becomes part of that cluster. If the read is not similar enough with any of the existing cluster centroids, it becomes the centroid of a new cluster. The similarity measure T is defined as a string similarity between the two nucleotide sequences that counts the number of character placements that they have in common and then divides that number by the length of the reads, with terminal characters excluded \cite{Edgar10}.

@@

We used a "vocabulary-based" feature extraction method. An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}. Step (i) is covered by our assembly and clustering process, while step (iii) is covered by performing classification with a standard SVM classifier based on the extracted feature vectors. In this case, we used svm-light \cite{joachims08}. Below, we describe our feature selection methods for step (ii), which we implemented in Python, as well as the rationale for using these methods.

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{./instance-labels.png}
\caption{This diagram illustrates why static instance labels are not sufficient for phenotype prediction. A patient with 6 of the blue microbe or 6 of the green microbe may be healthy, while a patient with 3 of each is sick. Static instance labels cannot capture this relationship. This is also explained by Amores \cite{amores13}.} \label{instance-labels}
\end{figure}

### 2016-10-29 latex/input/related.tex
results section comparison to mRMR+SVM updated
incorporated most of reviewer suggestions in paper

The Multiple Instance paradigm fits phenotype prediction well, since we have a set of labeled patients containing unlabeled sequence reads, and we would like to predict both the patient phenotype and which reads are indicative of that phenotype. Despite the recent developments in MIL and its potential utility in phenotype prediction, we have not found any literature that specifically applies MIL to classifying patient phenotype based on metagenomic data.

@@

One of the main purposes of assembly is to determine the whole genomes of microbial species, the vast majority of which have not or cannot be laboratory cultured, from sequencing reads \cite{zerbino08}. Even if complete genomes cannot be assembled, combining reads into larger contigs can still make them much more useful for clustering and classification, because the contigs will contain more phylogenetic and functional information than short reads. This is because short reads of less than 1000 base pairs constitute only a tiny fraction of microbial genomes, which are usually hundreds of thousands to millions of base pairs, making it difficult to ascertain much about the phylogeny of individual reads. Many modern sequence reads are produced by Next-Generation and High-Throughput Sequencers, which usually produce these short reads. Metagenomics additionally poses its own set of challenges, due to very large datasets and lack of knowledge about how many species are present and in what relative abundances \cite{namiki12}. Thus, metagenome assembly is a relatively new and challenging field. Some popular single genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, ABySS \cite{simpson09}, and Ray Meta \cite{boisvert12}. Here we use SOAPdenovo2 because it was the assembler used by Qin et al. in their metagenomics study \cite{qin041012}.
%There are variations on how exactly de Bruijn graphs are represented and implemented, but Figure \ref{de-bruijn} illustrates the general process.

>>>>

The Multiple Instance paradigm fits phenotype prediction well, since we have a set of labeled patients containing unlabeled sequence reads, and we would like to predict both the patient phenotype and which reads are indicative of that phenotype. Despite the recent developments in MIL and its potential utility in phenotype prediction, we have not found any literature that specifically applies MIL to classifying patient phenotype based on metagenomic data. We present our MIL-based feature extraction method in section \ref{feature-extraction}.

@@

One of the main purposes of assembly is to determine the whole genomes of microbial species, the vast majority of which have not or cannot be laboratory cultured, from sequencing reads \cite{zerbino08}. Even if complete genomes cannot be assembled, combining reads into larger contigs can still make them much more useful for clustering and classification, because the contigs will contain more phylogenetic and functional information than short reads. This is because short reads of less than 1000 base pairs constitute only a tiny fraction of microbial genomes, which are usually hundreds of thousands to millions of base pairs, making it difficult to ascertain much about the phylogeny of individual reads. Many modern sequence reads are produced by Next-Generation and High-Throughput Sequencers, which usually produce these short reads. Metagenomics poses its own set of challenges, due to large datasets and lack of knowledge about how many species are present and in what abundances \cite{namiki12}. Thus, metagenome assembly is a new and challenging field. Some popular single genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, and Ray Meta \cite{boisvert12}.
%There are variations on how exactly de Bruijn graphs are represented and implemented, but Figure \ref{de-bruijn} illustrates the general process.
